---
title: "Parametric survival models"
author: "Steve Simon"
output: 
  powerpoint_presentation
---

## Supplemental material

This presentation provides some interesting supplemental material for the module 5 lecture.

Here's a quick review of the likelihood ratio test, as it applies to the exponential accelerated time model.

First, let's fit a null model.

```{r load-libraries, echo=FALSE}
options(width=50)
library(knitr)
knitr::opts_chunk$set(echo=FALSE)
library(broom)
library(dplyr, warn.conflicts=FALSE)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr, warn.conflicts=FALSE, quietly=TRUE)
```

```{r null-model}
fn <- "../../data/whas100.RData"
load(fn)
null_model <- survreg(
  Surv(time_yrs, fstat=="Dead")~1,
  data=whas100,
  dist="exponential")
```

## Some summary statistics

```{r}
n0 <- sum(whas100$fstat=="Alive")
n1 <- sum(whas100$fstat=="Dead")
tbar <- round(mean(whas100$time_yrs), 3)
b0 <- round(log(tbar*(n0+n1)/n1), 3)
```

The intercept in the null model is `r b0`. That implies an overall hazard function of exp(`r b0`) = `r round(exp(b0), 2)`. There are `r n1` deaths and `r n0` censored observations. The average time of all patients (ignoring censoring) is `r tbar`.

## Mathematical details

How do you arrive at this mathematically. For the null model, the density is

$f(t) = \frac{1}{e^{\beta_0}} e^{-t/e^{\beta_0}}$

and the survival curve is

$S(t)=e^{-t/e^{\beta_0}}$ 

## Mathematical details

The likelihood for any parametric regression model is

$L(\beta)=\Pi_i f(t_i)^{c_i} S(t_i)^{1-c_i}$

Note that for deaths, you use the density, but for censored observations, you use the survival function, which can be thought of as the "average" density from time $t_i$ to infinity.

## Mathematical details

Let $n_0$ reprensent the number of censored observations and $n_1$ the number of deaths. Then

$L(\beta_0) = \Pi_1 \frac{1}{e^{\beta_0}} e^{-t/e^{\beta_0}} \Pi_0 e^{-t/e^{\beta_0}}$

Where $\Pi_1$ represents the product across all deaths and $\Pi_0$ represents the product across all censored observations. This simplifies to 

$L(\beta_0) = \frac{1}{e^{n_1 \beta_0}}  e^{-(n_0+n_1) \bar{t}/e^{\beta_0}}$

## Mathematical details

The log likelihood is

$l(\beta_0)= -n_1 \beta_0 - (n_0+n_1) \bar{t} e^{-\beta_0}$

To find the maximum, take the derivative and set it equal to zero.

$\frac{\partial l}{\partial \beta_0} = -n_1 + (n_0+n_1)\bar{t} e^{-\beta_0} = 0$

## Mathematical details

$(n_0+n_1)\bar{t} e^{-\hat{\beta_0}} = n_1$

$e^{\hat{\beta_0}} = \frac{n_0+n_1}{n_1} \bar{t}$

$\hat{\beta_0} = log\left(\frac{n_0+n_1}{n_1} \bar{t}\right)$

For the WHAS100 data set, the estimate of $\hat{\beta_0}$ is

log(((`r n0`+`r n1`) / `r n1`)  `r tbar`) = `r b0`.

## Mathematical details

The second derivative of the log likelihood is 

$\frac{\partial^2 l}{(\partial \beta_0)^2} = -(n_0+n_1)\bar{t} e^{-\beta_0}$

If you compute this negative of this second derivative at the maximum likelihood estimate, $\hat{\beta_0}$, you get the information matrix.

## Mathematical details

$I(\hat{\beta_0}) = (n_0+n_1)\bar{t} e^{-log\left(\frac{n_0+n_1}{n_1} \bar{t}\right)}$

$= (n_0+n_1)\bar{t} \left(\frac{n_1}{(n_0+n_1)\bar{t}} \right)$

$= n_1$

## Standard error

The standard error of the maximum likelihood estimate is 

$\frac{1}{\sqrt{I(\hat{\beta_0})}{}}$ 

which is simply 

$\frac{1}{\sqrt{n_1}{}}$

## Null calculations

```{r null-calculations, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
f <- function(t, b) {
  return(-b-t/exp(b))
}
s <- function(t, b) {
  return(-t/exp(b))
}
li <- function(t, c, b0) {
  # censored
  if (c!="Dead") {return(s(t, b0))}
  # death
  if (c=="Dead")  {return(f(t, b0))}
}
n0 <- 99
b0 <- seq(0, 2*2.09, length=n0)
l <- rep(0, n0)
for (i0 in 1:n0) {
  for (k in 1:100) {
    l[i0] <- l[i0] +
      li(whas100$time_yrs[k], whas100$fstat[k], b0[i0])
  }
}
m <- which.max(l)
data.frame(b0=b0, l=l) %>%
  ggplot(aes(b0, l)) +
  geom_line() +
  geom_label(aes(x=b0[m], y=l[m], label=round(l[m], 1)), size=2) +
  geom_label(aes(x=b0[1], y=l[1], label=round(l[1], 1)), size=2)
```

## Parametric model with gender

Now let's fit a model with gender as an independent variable. The survival curve is

$S(t)=e^{-t/e^{\beta_0}}$ 

for males and

$S(t)=e^{-t/e^{\beta_0+\beta_1}}$ 

for females.

## Parametric model with gender

The density functions for males and females are

$f(t) = \frac{1}{e^{\beta_0}} e^{-t/e^{\beta_0}}$

$f(t) = \frac{1}{e^{\beta_0+\beta_1}} e^{-t/e^{\beta_0+\beta_1}}$

## Parametric model with gender

The likelihood for any parametric regression model is

$L(\beta)=\Pi_i f(t_i)^{c_i} S(t_i)^{1-c_i}$

For this particular model, partition the observations into four sets: female censored ($F_0$), female deaths ($F_1$), male censored ($M_0$), and male deaths ($M_1$), with $f_0$, $f_1$, $m_0$, and $m_1$ observations in each group.

## Parametric model with gender

Then the likelihood becomes

$L(\beta_0, \beta_1) = \Pi_{F_0}  \Pi_{F_1} \Pi_{M_0} \Pi_{M_1}$

where

$\Pi_{F_0} = \Pi_{F_0} e^{-t_i/e^{\beta_0+\beta_1}}$

$\Pi_{F_1} = \Pi_{F_1}\frac{1}{e^{\beta_0+\beta_1}} e^{-t_i/e^{\beta_0+\beta_1}}$

$\Pi_{M_0} = \Pi_{M_0} e^{-t_i/e^{\beta_0}}$

$\Pi_{M_1} = \Pi_{M_1}\frac{1}{e^{\beta_0}} e^{-t_i/e^{\beta_0}}$

## Parametric model with gender

$L(\beta_0, \beta_1)=\frac{1}{e^{(f_1+m_1)\beta_0+f_1\beta_1}} e^{-\frac{(f_0+f_1)\bar{t}_F}{e^{\beta_0+\beta_1}} -\frac{(m_0+m_1)\bar{t}_M}{e^{\beta_0}}}$

## Parametric model with gender

Take the log of both sides to get a log likelihood.

$l(\beta0, \beta1)=-(f_1+m_1)\beta_0-f_1\beta_1 - (f_0+f_1)\bar{t}_F e^{-\beta_0-\beta_1} - (m_0+m_1)\bar{t}_M e^{-\beta_0}$

## Parametric model with gender

To find the maximum, take the two partial derivates and set them equal to zero.

$\frac{\partial l}{\partial \beta_0}=-(f_1+m_1) + (f_0+f_1)\bar{t}_F e^{-\beta_0-\beta_1} + (m_0+m_1)\bar{t}_M e^{-\beta_0}=0$

$\frac{\partial l}{\partial \beta_1}=-f_1 + (f_0+f_1)\bar{t}_F e^{-\beta_0-\beta_1}=0$

## Parametric model with gender

The solution to these two equations is 

$\hat\beta_0=log(\frac{m_0+m_1}{m_1}\bar{t}_M)$

$\hat\beta_1=log(\frac{f_0+f_1}{f_1}\bar{t}_F - \frac{m_0+m_1}{m_1}\bar{t}_M)$

## Parametric model with gender

```{r gender-model}
gender_model <- survreg(
  Surv(time_yrs, fstat=="Dead")~gender,
  data=whas100,
  dist="exponential")
tidy(gender_model)
m0 <- (whas100$gender=="Male" & whas100$fstat=="Alive")
m1 <- (whas100$gender=="Male" & whas100$fstat=="Dead")
f0 <- (whas100$gender=="Female" & whas100$fstat=="Alive")
f1 <- (whas100$gender=="Female" & whas100$fstat=="Dead")
tm <- round(mean(whas100$time_yrs[m0|m1]), 2)
tf <- round(mean(whas100$time_yrs[f0|f1]), 2)
```

## Calculate the parameters yourself

m0 = `r sum(m0)`, m1 = `r sum(m1)`, f0 = `r sum(f0)`, f1 = `r sum(f1)`

$\bar{t}_M$ = `r tm`, $\bar{t}_F$ = `r tf`, 

## Parametric model with gender

You can calculate the likelihoods yourself.

$L(\beta)=\Pi_i f(t_i, X_i, \beta)^{c_i} S(t_i, X_i, \beta)^{1-c_i}$

```{r create-function}
li <- function(t, c, g, b0, b1) {
  # male, censored
  if (c!="Dead" & g!="Female") {return(s(t, b0))}
  # male, death
  if (c=="Dead" & g!="Female")  {return(f(t, b0))}
  # female, censored
  if (c!="Dead" & g=="Female") {return(s(t, b0+b1))}
  # female, death
  if (c=="Dead" & g=="Female")  {return(f(t, b0+b1))}
}
```

## Draw the likelihood surface

```{r surface, fig.width=4.5, fig.height=2.5}
par(mar=c(0.6, 0.6, 0.6, 0.6))
n0 <- 49
n1 <- 49
b0 <- seq(2.318-0.602, 2.318+0.602, length=n0)
b1 <- seq(-2*0.602, 0, length=n1)
l <- matrix(0, n0, n1)
for (i0 in 1:n0) {
  for (i1 in 1:n1) {
    for (k in 1:100) {
      l[i0, i1] <- l[i0, i1] +
            li(whas100$time_yrs[k], 
               whas100$fstat[k],
               whas100$gender[k],
               b0[i0], b1[i1])
    }
  }
}
persp(b0, b1, l, theta=45, phi=30)
```

## Draw contour plot.

```{r contour, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
b0 <- seq(0, 2.318+0.602, length=n0)
b1 <- seq(-2*0.602, 0, length=n1)
l <- matrix(0, n0, n1)
for (i0 in 1:n0) {
  for (i1 in 1:n1) {
    for (k in 1:100) {
      l[i0, i1] <- l[i0, i1] +
            li(whas100$time_yrs[k], 
               whas100$fstat[k],
               whas100$gender[k],
               b0[i0], b1[i1])
    }
  }
}
lv <- -c(156:160, 162, 165, 170, 180, 190, 200, 250, 300, 350, 400, 600, 800)
contour(b0, b1, l, levels=rev(lv))
# abline(h=-0.602, lty="dotted")
# abline(v=2.318, lty="dotted")
m <- b0[which.max(l[ , n1])]
text(2.318, -0.602, round(max(l), 1), col="red", cex=0.5)
text(m, 0, round(max(l[ , n1]), 1), col="red", cex=0.5)
text(0, 0, round(l[1, n1], 1), col="red", cex=0.5)
```
