---
title: "weibull-simulation5"
author: "Steve Simon"
date: "October 19, 2018"
output: html_document
---

This program shows the bare bones approach to simulation.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Here's a hypothetical scenario. You collect data on 50 patients in each of two groups (total of 100 patients). In the both groups, you think that the distribution of survival times follows a Weibull distribution with shape parameter 1.2 (implying a slightly increasing hazard over time). The scale parameters for the two groups are 2 and 5, respectively, implying a decelleration of time in the second group (fewer events). With this sized difference, what would your power be?

You have to account for censoring, of course. There are two ways that an observation could be censored. Either the event occurs after the maximum follow-up time of 5 years, or the patients drop out early. The dropout process is an exponential distribution with a rate of 0.25.

```{r generate}
n1 <- 50
n2 <- 50
n_reps <- 1000
gp <- rep(rep(0:1, c(n1, n2)), n_reps)
rp <- rep(1:n_reps, each=n1+n2)
event_time <- 
  rweibull(n_reps*(n1+n2), shape=1.2, scale=2+3*gp)
dropout_time <- rexp(n_reps*(n1+n2), rate=0.25)
time <- pmin(event_time, dropout_time, 5)
censor_at_5 <- event_time > 5 & dropout_time > 5
early_dropout <- dropout_time < event_time & event_time <=5
mean(censor_at_5)
mean(early_dropout)
cens <- as.numeric(!censor_at_5 & !early_dropout)
df <- data.frame(time, cens, gp, rp)
event_counts <- tapply(df$cens, list(df$rp, df$gp), sum)
head(event_counts, 20)
```

Run the analyses.

```{r run, error=TRUE}
output <- NULL
save(df, file="tst1.RData")
for (i in 1:n_reps) {
  sb <- filter(df, rp==i)
  surv_data <- Surv(sb$time, sb$cens)
  m1 <- sum(sb$cens==1 & sb$gp==0)
  m2 <- sum(sb$cens==1 & sb$gp==1)
  weibull_model <- 
    survreg(surv_data~gp, 
      data=sb, dist="weibull")
  if (weibull_model$iter==30) {next}
  weibull_model %>%
    tidy %>%
    filter(term=="gp") %>%
    select(term, estimate, p.value) %>%
    bind_rows(output) -> output
}
n_reps - dim(output)[1] # replications with no events
head(output, n=20)
mean(output$p.value < 0.05)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("../../data/weibull-simulation5.RData")
```